---
title: "Operations Research & Data Analytics - Final Project"
author: "Joey Clancy"
date: "12/2/2021"
output:
  html_document:
    code_download: yes
    toc: yes
    toc_float: yes
    number_sections: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---

# Project Goal

---

# Data Brief & Project Notes

## About the data

## Addressing missing values

---


# Setup

```{r clear-environment, echo = FALSE}
# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)

```

```{r message=FALSE, warning=FALSE, echo=FALSE}
# Import packages
library(tidyverse)
library(gridExtra)
library(scales)
library(caTools)
#library(mice)
library(data.table)
#library(doParallel)
library(kableExtra)
library(stringr)
library(fixest)
library(modelsummary)
library(lubridate)
```

# Data Preprocessing

```{r warning=FALSE}
# Load data
covid <- read.csv("fullCOVIDtable.csv", sep = ";")
covid <- covid[,-1]
```

## Original Data Set

```{r}
# Change position of some columns
covid <- covid %>% relocate(CONTINENT, .after = X20FLIGHTS)
covid <- covid %>% relocate(COUNTRY, .after = CONTINENT)
covid <- covid %>% relocate(DATE, .after = COUNTRY)
# Drop electrical information columns as they will not be used and are largely NA
covid <- covid[,1:37]

# Remove regional data
covid <- covid[!grepl("-", covid$COUNTRY),]

# Create df with factors for visualizations
covidFact <- covid
covidFact <- covidFact[!grepl("-", covidFact$COUNTRY),]

# Convert vars to factor type
cols <- c("COUNTRY", "CONTINENT", "SCHOOL", "WORK", "EVENTS", "GATHERINGS",
          "TRANSPORTATION", "ATHOME", "NATIONAL", "INTERNATIONAL", "INFORMATION",
          "TESTING", "TRACING")
covidFact[cols] <- lapply(covid[cols], as.factor)

# Convert date column to date type
covid$DATE <- as.Date(covid$DATE)
covidFact$DATE <- as.Date(covid$DATE)
```

## Imputing Missing Values

```{r}
# Deal with missing/infinite values
covid <- covid %>% 
    mutate_if(is.numeric, ~ replace(., is.infinite(.), NA))
# Create 3 imputed data sets with mice package
#covidImp <- mice(covid, m=3, maxit = 5, method = 'rf', seed = 777)    # Impute missing data using random forests
```

## New Completed Data Creation

```{r}
# Combines 3 imputed data sets for pooling. To be used in statistical EDA later on. 
#completedData <- complete(covidImp, action = "long")
```

## Export Completed Data Set

```{r}
# Export completed data to csv in order to avoid repeating previous steps
#write.csv(completedData, file = "covidImp.csv")
```

## Load Imputed Data

```{r}
# Import completed data
completedData <- read.csv("covidImp.csv")
# Get rid of row IDs
completedData <- completedData[,c(-1, -4)]
```

## Wrangle New Data

```{r}
# Convert vars to factor type
# cols <- c("COUNTRY", "CONTINENT", "SCHOOL", "WORK", "EVENTS", "GATHERINGS",
#          "TRANSPORTATION", "ATHOME", "NATIONAL", "INTERNATIONAL", "INFORMATION",
#          "TESTING", "TRACING")

# completedData[cols] <- lapply(completedData[cols], as.factor)

# Convert date column to date type
completedData$DATE <- as.Date(completedData$DATE)

# Get rid of regional data
completedData <- completedData[!grepl("-", completedData$COUNTRY),]

```


# Univariate Non-Graphical

```{r}
# Summary of completed data
summary(completedData)
```

```{r}
# Structure of completed data
str(completedData)
```

```{r}
summary(covid)
```


# Univariate Graphical

## Log Distribution Comparisons

```{r}
# The distributions are expected to be similar between the two data sets
grid.arrange(
    ggplot(data = completedData, mapping = aes(y = log(Cases))) +
      geom_density() + 
        theme_classic()+
      coord_flip() +
      labs(title = "Log Cases - Imputed Data Set"),
    
    ggplot(data = covid, mapping = aes(y = log(Cases))) +
      geom_density() + 
        theme_classic() +
      coord_flip() +
      labs(title = "Log Cases - Original Data Set")
)
```

## SCHOOL

```{r}
completedData %>% 
  ggplot(mapping = aes(x = SCHOOL)) +
  geom_bar()
```

## WORK

```{r}
completedData %>% 
  ggplot(mapping = aes(x = WORK)) +
  geom_bar()
```

## EVENTS

```{r}
completedData %>% 
  ggplot(mapping = aes(x = EVENTS)) +
  geom_bar()
```

## GATHERINGS

```{r}
completedData %>% 
  ggplot(mapping = aes(x = GATHERINGS)) +
  geom_bar()
```

## TRANSPORTATION

```{r}
completedData %>% 
  ggplot(mapping = aes(x = TRANSPORTATION)) +
  geom_bar()
```

## ATHOME

```{r}
completedData %>% 
  ggplot(mapping = aes(x = ATHOME)) +
  geom_bar()
```

## NATIONAL

```{r}
completedData %>% 
  ggplot(mapping = aes(x = NATIONAL)) +
  geom_bar()
```

## INTERNATIONAL

```{r}
completedData %>% 
  ggplot(mapping = aes(x = INTERNATIONAL)) +
  geom_bar()
```

## INFORMATION

```{r}
completedData %>% 
  ggplot(mapping = aes(x = INFORMATION)) +
  geom_bar()
```

## TESTING

```{r}
completedData %>% 
  ggplot(mapping = aes(x = TESTING)) +
  geom_bar()
```

## TRACING

```{r}
completedData %>% 
  ggplot(mapping = aes(x = TRACING)) +
  geom_bar()
```

```{r}
# Function to create density and box plot pair
densityBox <- function(vbl) {
  grid.arrange(
    ggplot(data = completedData, mapping = aes(x = {{vbl}})) +
      geom_density() + 
        theme_classic(),
    
    ggplot(data = completedData, mapping = aes(x = 0)) +
      geom_boxplot(mapping = aes(y = {{vbl}})) +
        coord_flip() + 
        theme_classic()
  )
}
```

## OLD

```{r}
densityBox(OLD)
```

## YOUNG

```{r}
densityBox(YOUNG)
```

## URBAN

```{r}
densityBox(URBAN)
```

## DENSITY

```{r}
densityBox(DENSITY)
```

## POPULATION

```{r}
densityBox(POPULATION)
```

## PHYSICIANS

```{r}
densityBox(PHYSICIANS)
```

## BEDS

```{r}
densityBox(BEDS)
```

## Cases

```{r}
densityBox(Cases)
```

## Deaths

```{r}
densityBox(Deaths)
```

## New.Cases

```{r}
densityBox(New.Cases)
```

## New.Deaths

```{r}
densityBox(New.Deaths)
```

## REFF

```{r}
densityBox(REFF)
```

## INFECTED

```{r}
densityBox(INFECTED)
```

## INFECTEDINCREASE

```{r}
densityBox(INFECTEDINCREASE)
```

## MORTALITY

```{r}
densityBox(MORTALITY)
```

## FATALITYINCREASE

```{r}
densityBox(FATALITYINCREASE)
```

## TEMPERATURE

```{r}
densityBox(TEMPERATURE)
```

## HUMIDITY

```{r}
densityBox(HUMIDITY)
```

## WIND

```{r}
densityBox(WIND)
```

## FLIGHTEVOLUTION

```{r}
densityBox(FLIGHTEVOLUTION)
```

## 19Flights

```{r}
densityBox(X19FLIGHTS)
```

## 20FLIGHTS

```{r}
densityBox(X20FLIGHTS)
```

# Multivariate Non-graphical

```{r}
# Get correlation matrix
cormat <- cor(completedData[, unlist(lapply(completedData, is.numeric))])
kable(cormat)
```

# Multivariate Graphical

## Correlation Heat Map

```{r}
# Melt the previous correlation matrix
melted_cormat <- melt(cormat)

# Visualize the correlations in a heat map
ggplot(data = melted_cormat, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Correlation Heat Map")
```

## US vs. World - New Cases

**Note**: The plot below shows the number of new cases in the US over the course of 2020, compared with the maximum number of new cases across all countries. This means that any point on the red line could have a different country associated with it. Put simply, the red line is made up of all countries that had the max number of new cases on any respective day of the year.

```{r message=FALSE, warning=FALSE}
# Using original data set
covid %>% 
  filter(New.Cases <= 800000) %>% # For some reason there was an 800000 outlier that appeared in the graph but not in the actual data, hence the filter
  select(DATE, New.Cases, COUNTRY) %>% 
  mutate(unitedStates = as.factor(ifelse(COUNTRY == "US", "US", "Other"))) %>% #changing 0,1 to no,yes
  group_by(DATE, unitedStates) %>% 
  summarize(globalMax = max(New.Cases)) %>% 
  ggplot(mapping = aes(x = DATE, y = globalMax/1000, color = unitedStates)) + #using median price 
  geom_line() +
  theme_minimal() +
  labs(title = "New US Cases vs. Max World New Cases (per day)",
       subtitle = "Dramatic increase in new US-based cases starting in early Winter",
       x = "Date",
       y = "New Cases (thousands)") + 
  gghighlight::gghighlight(max(globalMax) > 150000)
```

# Statistical EDA

##### Notes

- The following is an examination of the effect of all forms of policy on the effective reproduction number.

- The effective reproduction number or Reff, is the average number of secondary infections caused by a single infected case, in a fully susceptible population.

- The following three experiments look at the United States, China, and India.

**There were some assumptions and/or flaws made while building these models:**

1. All countries enact policies at the same time (obviously not true, but necessary in order to create control and treatment groups).

2. The effects of the various levels within a policy category are negligible.

3. Treatment consists of the enactment of policies across *all* categories at some level above zero.

4. Policies stay active, so no policy reverts back to zero (inactive)

## Difference in Differences Setup

```{r}
# Retrieve columns of interest
colListDiD <- c("SCHOOL", "WORK", "EVENTS", "GATHERINGS",
          "TRANSPORTATION", "ATHOME", "NATIONAL", "INTERNATIONAL", 
          "INFORMATION", "TESTING", "TRACING", "Cases", "Deaths", "New.Cases",
          "New.Deaths", "REFF", "INFECTED", "INFECTEDINCREASE", "MORTALITY", 
          "FATALITYINCREASE", "COUNTRY", "DATE")
# Create new data set from the completed set
did <- covid[,colListDiD]


# Create month column, use as time period measure
did$DATE <- as.Date(did$DATE, format = "%Y-%m-%d")
did[, "MONTH"] <- format(did[,"DATE"], "%m")

# Create Day column
#did[, "DAY"] <- format(did[,"DATE"], "%d")

#did$DAY <- as.numeric(did$DAY)
did$MONTH <- as.numeric(did$MONTH)

did[, 1:20][is.na(did[, 1:20])] <- 0
```

```{r}
did <- did %>% 
  mutate(allPolicies = if_else(rowSums(did[,1:11] == 0) > 0, 0, 1))   
# If a row had all policies implemented at some level, set allPolicies == 1
# Likewise, if there were only some policies implemented, set allPolicies == 0
```

```{r}
# Control group is made up of countries with the lowest totals (equal to zero) of the allPolicies variable
controlGroupID <- did %>% 
  filter(COUNTRY != "") %>% 
  group_by(COUNTRY) %>% 
  summarize(allPolicyCheck = sum(allPolicies))

controlGroupID <- controlGroupID[order(controlGroupID$allPolicyCheck),]

controlGroup <- controlGroupID[1:73, 1]

# write.csv(controlGroup, file = "controlG.csv", row.names = FALSE)
## Can then copy and paste country names below

controlGroup <- c("Andorra","Angola","Anguilla","Antigua and Barbuda","Armenia","Aruba","Belarus","Belgium","Benin","Bonaire, Sint Eustatius and Saba",
"Bosnia and Herzegovina","British Virgin Islands","Brunei","Bulgaria","Burma","Burundi","Cambodia","Cameroon","Canada","Cayman Islands",
"Central African Republic","Channel Islands","Comoros","Curacao","Czechia","Dominica","Equatorial Guinea","Estonia","Falkland Islands (Malvinas)",
"Finland","French Guiana","French Polynesia","Grenada","Guadeloupe","Iceland","Isle of Man","Japan","Korea, South","Latvia","Liechtenstein",
"Malawi","Maldives","Mali","Martinique","Mauritania","Montenegro","Montserrat","Namibia","New Caledonia","Nicaragua","Niger","North Macedonia","Norway",
"Saint Barthelemy","Saint Kitts and Nevis","Saint Lucia","Saint Pierre and Miquelon","Saint Vincent and the Grenadines","Sao Tome and Principe","Senegal",
"Singapore","Sint Maarten","Somalia","St Martin","Switzerland","Taiwan*","Tajikistan","Tanzania","Turks and Caicos Islands","Venezuela","Western Sahara",
"Yemen","Zambia","US","China","India")
```

```{r}
did <- subset(did, COUNTRY %in% controlGroup)
```

### United States

```{r}
# Creates treatment and control groups
did <- did %>% 
  mutate(group = if_else(COUNTRY == "US", 1, 0)) %>% 
  # Use March 19th as cutoff (date when all varieties of policy had been enacted)
  mutate(after = if_else(DATE >= "2020-03-19", 1, 0)) %>% 
  mutate(treated = after * (group==1))
```

```{r}
# Get average REFF both before and after the treatment for both groups
means <- did %>% 
  group_by(group, after) %>% 
  summarize(AvgReff = mean(REFF))
means
```

```{r}
# Control for changes over time and differences in groups
bef.aft.untreated <- filter(means, after == 1, group == 0)$AvgReff - filter(means, after == 0, group == 0)$AvgReff
bef.aft.treated <- filter(means, after == 1, group == 1)$AvgReff - filter(means, after == 0, group == 1)$AvgReff

bef.aft.treated
bef.aft.untreated

# Retrieve DID Estimator by taking out the difference we see in the control group
bef.aft.treated - bef.aft.untreated
```

```{r warning=FALSE}
m1 <- feols(REFF ~ treated | group + after, data = did)
msummary(m1, stars = TRUE, gof_omit = 'AIC|BIC|Lik|Adj|Pseudo')
```

### China

```{r}
# Creates treatment and control groups
did <- did %>% 
  mutate(group = if_else(COUNTRY == "China", 1, 0)) %>% 
  # Use March 19th as cutoff (date when all varieties of policy had been enacted)
  mutate(after = if_else(DATE >= "2020-02-25", 1, 0)) %>% 
  mutate(treated = after * (group==1))
```

```{r}
# Get average REFF both before and after the treatment for both groups
means <- did %>% 
  group_by(group, after) %>% 
  summarize(AvgReff = mean(REFF))
means
```

```{r}
# Control for changes over time and differences in groups
bef.aft.untreated <- filter(means, after == 1, group == 0)$AvgReff - filter(means, after == 0, group == 0)$AvgReff
bef.aft.treated <- filter(means, after == 1, group == 1)$AvgReff - filter(means, after == 0, group == 1)$AvgReff

bef.aft.treated
bef.aft.untreated

# Retrieve DID Estimator by taking out the difference we see in the control group
bef.aft.treated - bef.aft.untreated
```

```{r}
m2 <- feols(REFF ~ treated | group + after, data = did)
msummary(m2, stars = TRUE, gof_omit = 'AIC|BIC|Lik|Adj|Pseudo')
```

### India

```{r}
# Creates treatment and control groups
did <- did %>% 
  mutate(group = if_else(COUNTRY == "India", 1, 0)) %>% 
  # Use March 19th as cutoff (date when all varieties of policy had been enacted)
  mutate(after = if_else(DATE >= "2020-03-20", 1, 0)) %>% 
  mutate(treated = after * (group==1))
```

```{r}
# Get average REFF both before and after the treatment for both groups
means <- did %>% 
  group_by(group, after) %>% 
  summarize(AvgReff = mean(REFF))
means
```

```{r}
# Control for changes over time and differences in groups
bef.aft.untreated <- filter(means, after == 1, group == 0)$AvgReff - filter(means, after == 0, group == 0)$AvgReff
bef.aft.treated <- filter(means, after == 1, group == 1)$AvgReff - filter(means, after == 0, group == 1)$AvgReff

bef.aft.treated
bef.aft.untreated

# Retrieve DID Estimator by taking out the difference we see in the control group
bef.aft.treated - bef.aft.untreated
```

```{r}
m3 <- feols(REFF ~ treated | group + after, data = did)
msummary(m3, stars = TRUE, gof_omit = 'AIC|BIC|Lik|Adj|Pseudo')
```






























